Targeted interrogation of mutational hotspots in rheumatoid arthritis

### 2024-03-11 scomatic/reports/targeted_mutation_calling.Rmd
fix targeted_mutation_calling and generate notebook of the results

---
title: "Targeted interrogation of mutational hotspots in rheumatoid arthritis"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: false
    code_folding: show
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
params:
  drivers: NULL
  hotspots: NULL
  bam_mappings: NULL
  cache_dir: './target_mutation_calling_cache/'
  rerun: true
---

```{r setup, include = F, message = F, warning = F, class.source = 'fold-hide'}
# wd <- ifelse(Sys.info()['nodename'] == "mib119104s", '~/Volumes/', '') |> paste0('/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/') ; setwd(wd)
# rmarkdown::render('reports/driver_coverage.Rmd', output_file = 'driver_coverage.html', output_dir = 'reports/', params = list(rerun = F))

# chunk options
knitr::opts_knit$set(root.dir = '../')
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  dpi = 300, 
  cache.path = params$cache_dir)

@@

```{r load_files}
list.files(pattern = '_reads.tsv') %>%
  purrr::set_names(., gsub('_reads.tsv', '', .)) %>%
  purrr::map(function(file) {
    reads <- readr::read_tsv(reads)
  })
```

```{r plot, fig.width = 20}
# plot positions
unique(sites$gene) %>%
  purrr::walk(function(i) {
    p <-
      sites %>%
      dplyr::filter(gene == i) %>%
      dplyr::left_join(refs, by = c('chr', 'pos')) %>%
      tidyr::pivot_longer(cols = dplyr::all_of(nucleotides),
                          values_to = 'nt_n_reads', names_to = 'nt') %>%
      dplyr::mutate(
        strandless_nt = 
          ifelse(nt %in% c('-', '_'), 'DEL', toupper(nt)),
        is_ref = ifelse(strandless_nt == ref, 'ref', 'alt'),
        strandless_nt = paste0(strandless_nt, '_', is_ref) %>% 
          forcats::fct_relevel(unique(c(paste0(toupper(nucleotides), '_ref'),
                                        paste0(toupper(nucleotides), '_alt'))))) %>%
      ggplot(aes(x = pos, y = nt_n_reads, fill = strandless_nt)) + 
      geom_vline(aes(xintercept = mut_pos)) +
      geom_col(position = 'stack') +
      theme_bw() +
      theme(axis.text.x = element_blank()) +
      ggh4x::facet_grid2(celltype ~ mut_pos, scales = 'free') +
      scale_fill_manual(values = c(A_ref = 'darkgreen', A_alt = 'lightgreen',
                                   C_ref = 'darkblue', C_alt = 'lightblue',
                                   G_ref = 'darkorange3', G_alt = 'peachpuff',
                                   T_ref = 'darkred', T_alt = 'salmon',
                                   INS_alt = 'black', DEL_alt = 'grey')) +
      ggtitle(i)
    print(p)
    })
```

# Mutational calls in drivers from the LCM-WES data

```{r mut_hotspots, eval = F}
# read in bam mappings, get celltype-split bams only
mappings <-
  readr::read_csv(params$bam_mappings) %>%
  dplyr::filter(!is.na(celltype)) %>%
  # temporarily remove problematic BRI-401 endo bam
  dplyr::filter(!(id == 'BRI_401' & celltype == 'Endothelial_cell'))

# read in drivers
drivers <-
  readr::read_tsv(params$drivers, col_names = c('chr', 'start', 'end', 'gene')) %>%
  dplyr::mutate(chr = gsub('chr', '', chr))

# read in mutations in drivers
hotspots <-
  readr::read_tsv(params$hotspots) %>%
  dplyr::mutate(mut_pos = pos,
                chr = gsub('chr', '', chr)) %>%
  dplyr::group_by(chr, mut_pos, gene, ref_cod) %>%
  # summarise each position
  dplyr::summarise(n = dplyr::n(),
                   id = paste(id, collapse = ','),
                   mut_cod = paste(mut_cod, collapse = ','),
                   impact = paste(impact, collapse = ',')) %>% 
  dplyr::group_by(dplyr::across(everything())) %>%
  # extend pileup window 20bp either side of mutated position
  dplyr::reframe(pos = (mut_pos - 20):(mut_pos + 20))

# save as a bed
hotspots %>%
  dplyr::transmute(chr = paste0('chr', chr), start = pos - 1, end = pos) %>%
  readr::write_tsv(
    'data/driver_genes/lcm_wes_mutations_20bp_window.bed',
    col_names = F)

# get ref allele at each position
refs <- xfun::cache_rds({
  refs <-
    system('bedtools getfasta -fi /lustre/scratch125/casm/team268im/al28/bed_ref/GRCh38_full_analysis_set_plus_decoy_hla_genome.fa -bed data/driver_genes/lcm_wes_mutations_20bp_window.bed -tab', intern = T) %>%
    paste(collapse = '\n') %>%
    readr::read_tsv(col_names = c('pos', 'ref')) %>%
    tidyr::separate_wider_delim(cols = pos, delim = stringr::regex(':|-'), 
                                names = c('chr', 'start', 'end')) %>%
    type.convert(as.is = T) %>%
    dplyr::transmute(chr = gsub('chr', '', chr), pos = end, ref) %>%
    dplyr::distinct()
  refs
}, file = 'refs.rds', rerun = params$rerun)
```

>>>>

---
title: "Targeted interrogation of mutational hotspots in rheumatoid arthritis"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: true
    code_folding: show
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
params:
  cache_dir: './target_mutation_calling_cache/'
  rerun: true
---

```{r setup, include = F, message = F, warning = F, class.source = 'fold-hide'}
# wd <- ifelse(Sys.info()['nodename'] == "mib119104s", '~/Volumes/', '') |> paste0('/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/') ; setwd(wd)

# chunk options
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  dpi = 300, 
  cache.path = params$cache_dir)

@@

```{r load_files}
# all possible nucleotides
nucleotides <- c("A", "T", "C", "G", "-", "INS", "a", "t", "c", "g", "_", "ins")

# strandless nucleotides - colours and order
nt_cols <-
  c('INS' = '#e6ab02',
    'DEL' = '#1c2d78',
    'A' = '#1b9e77', 
    'C' = '#d95f02', 
    'G' = '#7570b3', 
    'T' = '#e7298a',
    'ref' = 'grey')

# recursites (to add detected mut)
recurmuts <-
  readr::read_tsv('recursites.tsv') %>%
  dplyr::transmute(
    chr,
    mut_pos = pos,
    mut_pos_lab = paste0(pos, '\ mut: ', mut)
  )

# reads
reads <-
  list.files(pattern = '_reads.tsv') %>%
  purrr::set_names(., gsub('_reads.tsv', '', .)) %>%
  purrr::map(function(file) {
    readr::read_tsv(file, show_col_types = F)
  }) %>%
  dplyr::bind_rows(.id = 'id') %>%
  tidyr::pivot_longer(cols = dplyr::all_of(nucleotides)) %>%
  dplyr::mutate(
    # colour in alt alleles
    status = dplyr::case_when(toupper(name) == ref ~ 'ref',
                              name %in% c('-', '_') ~ 'DEL',
                              TRUE ~ toupper(name)) %>%
      forcats::fct_relevel(names(nt_cols))) %>%
  # add detected mutations
  dplyr::left_join(recurmuts, by = c('chr', 'mut_pos')) %>%
  split(., .$gene)
```

## Mutations

First, we plot the rates of reference and non-reference alleles at mutation windows in all genes. Then, we do the same for substitutions only (ignoring indels).

```{r plot}
# function: create plot list
plot_positions <- function(reads, include_indels = T) {
  p <-
    names(reads) %>%
    purrr::set_names() %>%
    purrr::map(function(gene) {

      p_dat <-
        reads[[gene]] %>%
        # optionally exclude indels
        dplyr::filter(include_indels == T |
                !name %in% c('-', '_', 'INS', 'ins')) %>%
        # only plot celltypes with any reads
        dplyr::group_by(celltype) %>%
        dplyr::filter(max(value) > 0) %>%
        dplyr::ungroup()

      if (nrow(p_dat) > 0) {

        # get ref alleles for x axis labels
        x_labs <-
          p_dat %>%
          dplyr::distinct(pos, ref) 
        x_labs_vec <- x_labs$ref
        names(x_labs_vec) <- x_labs$pos

        g_title <- paste0(
          gene, ' w/', 
          ifelse(include_indels == T, '', 'o'), 
          ' indels')

        # get plot dims
        g_p_dims <-
          c(min(max(5, dplyr::n_distinct(p_dat$mut_pos) * 2 + 2), 50),
            min(max(5, dplyr::n_distinct(p_dat$celltype) * 1.7 + 2), 50))

        # get plot
        g_p <-
          p_dat %>%
          dplyr::mutate(
            # make sure ordering of positions is maintained
            x_pos = factor(as.character(pos), 
                           levels = unique(as.character(sort(pos)))),
            x_mut_pos = dplyr::case_when(pos == mut_pos ~ x_pos, TRUE ~ NA)) %>%
          # plot
          ggplot(aes(x = x_pos, y = value, fill = status)) +
          geom_col(position = 'stack') +
          geom_vline(aes(xintercept = x_mut_pos)) +
          theme_classic() +
          scale_x_discrete(labels = x_labs_vec) +
          scale_fill_manual(values = nt_cols) +
          ggh4x::facet_grid2(celltype ~ mut_pos_lab, scales = 'free') +
          labs(title = g_title, x = 'pos (ref shown)')

        # return
        list(plot = g_p, dims = g_p_dims)

      }
    })

  # return
  p
}

# generate plot lists
p_w_indels <- plot_positions(reads)
p_wo_indels <- plot_positions(reads, F)
```

```{r, results = 'asis'}
chunks <- c()

for (gene in names(reads)) {

  chunks <- c(chunks, paste0('\n### ', gene))

  knit_plot <- c(
    '',
    paste0('```{r ', gene,'_w_indels, fig.dim = p_w_indels[["', gene, '"]]$dims, warning = F, message = F, echo = F}'),
    paste0('print(p_w_indels[["', gene, '"]]$plot)'),
    '```',
    '',
    paste0('```{r ', gene,'_wo_indels, fig.dim = p_wo_indels[["', gene, '"]]$dims, warning = F, message = F, echo = F}'),
    paste0('print(p_wo_indels[["', gene, '"]]$plot)'),
    '```',
    ''
  )

  knitted_plot <- knitr::knit_child(text = knit_plot, envir = environment(), quiet = T)
  dev_off_if()

  chunks <- c(chunks, knitted_plot)

}

cat(unlist(chunks), sep = '\n')
```

## Mutations types, by patient across genes

```{r muts_by_patient, fig.width = 10}
reads %>%
  dplyr::bind_rows() %>%
  ggplot(aes(x = id, y = value, fill = status)) +
  geom_col() +
  scale_fill_manual(values = nt_cols) + 
  theme_classic() +
  scale_x_discrete(guide = guide_axis(angle = -90))
```
